Gusto: a toolkit for compatible finite element dynamical cores

### 2025-08-05 source/paper.tex

\subsubsection{Transport of $\mathbb{V}^{HDiv}$ fields}
All three of the equation sets described in section \ref{} include a
prognostic equation for the velocity field
$\MM{u}\in\mathbb{V}^{HDiv}$ that contains a nonlinear transport term
$\MM{u}\cdot\nabla\MM{u}$. This term can be discretised directly using
a vector version of the scheme in section ** or, as in \citet{} it can
be replaced by the vector invariant form and then discretised. In
Gusto we offer both options and we present the vector invariant scheme
below.

The schemes described above do not preserve monotonicity unlike the
continuous equation, generating spurious over and undershoots. When
monotonicity is important, for example, to avoid generating spurious
and unphysical negative values of moisture variables, we can apply a
limiter as part of the transport scheme. These limiters are described
in \ref{}.

It is common in weather and climate models for the horizontal and
vertical directions to be discretised differently due to the different
scales of motion in each direction. Typically the vertical part of the
transport term is treated implicitly so that the timestep is not
restricted by the vertical Courant number. We provide the option to
split the transport terms into horizontal and vertical components, as
described in \ref{}.

@@

\section{Results}
\label{sec: results}
The aim of this section is to present a selection of results that
demonstrate the tools available in Gusto, such as the choice of grids,
HDiv element families, order of finite element spaces, different
spatial discretisation options, timestepping schemes and physics. The
numerical properties of these configurations (convergence, stability,
accuracy and efficiency) have either been demonstrated elsewhere, or
will be investigated more thoroughly in future publications.

To include:
\begin{itemize}
\item terminator toy (Tim)
\item shallow water on different grids and with different HDiv
  families (Nell) e.g. Williamson 5 cubed sphere PV and Galewsky
  icosahedral sphere PV, high res (ref 6)
\item convergence with different RK schemes / multistage schemes using
  Williamson 2 or 6, L2 and Linf errors for D and u (2 plots, one L2,
  one Linf) (Alex)
\item vertical slice Euler - Straka, different orders, illustrates
  recovery for diffusion (Tom)
\item 3D moist bubble with rain (Tom/Alex)
\item spherical gravity wave, Boussinesq (Jemma)
\item spherical baroclinic wave, Euler - whatever people usually plot
  (horizontal slice pressure contours, temperature shading) (Dan)
\item Held Suarez??
\end{itemize}

\subsection{Terminator `toy' chemistry}
This test combines transport and a `toy' chemistry model to
demonstrate physics-dynamics coupling in Gusto (where here chemistry
is the `physics').

\subsection{Shallow water}
The standard suite of shallow water test cases originally proposed in
\citet{williamson1992standard} provides a common starting point for
testing new discretisations. This test suite is usually augmented by
the unstable barotropic jet described in
\citet{galewsky2004initial}. Here we present convergence results for a
variety of timestepping schemes applied to the steady state
geostrophic balanced flow (test case 2 from
\citet{williamson1992standard}) and a demonstration of two different
grids and HDiv element families for shallow water flow over a mountain
(test case 5 from \citet{williamson1992standard}) and the unstable
barotropic jet \citep{galewsky2004initial}.

\subsubsection{Convergence results}

\subsubsection{Grids and finite element families}

\subsection{Compressible Boussinesq equations}

\subsection{Compressible Euler equations}

\subsubsection{Dense bubble in a vertical slice}

\subsubsection{3D moist rising bubble}

\subsubsection{Baroclinic wave on the sphere}

\subsubsection{Held Suarez}

>>>>

\subsubsection{Transport of $\mathbb{V}^{HDiv}$ fields}

These schemes do not preserve the monotonicity of the continuous
equation, generating spurious over and undershoots. When monotonicity
is important, for example, to avoid generating spurious and unphysical
negative values of moisture variables, we can apply a limiter as part
of the transport scheme.

\JScomment{I think that we should at least mention that we can do horizonal and vertical splitting for stability - maybe don't need to write out all of the forms though!}

@@

\section{Results}
\label{sec: results}
The aim of this section is to present a selection of results that
showcase the tools available in Gusto, such as the choice of grids,
HDiv element families, order of finite element spaces, different
spatial discretisation options and timestepping schemes.
As these options can be combined in so many ways, it is only possible to present results from a tiny fraction of the configurations.
We have chosen to use a selection from the hierarchy of standard test cases that span the stages of development of dynamical cores.
Although we do include some results here that demonstrate the numerical properties of certain configurations, the purpose of this section is not to provide a rigorous verification of any particular choice of discretisation. Such results are demonstrated elsewhere, for instance \cite{bendall2020compatible} or \cite{hartney2025exploring}.
The settings used for each result from this section are detailed in Appendix \ref{sec:appendix}.

\subsection{Terminator Toy chemistry}
The first results use the ``Terminator Toy'' test of \cite{lauritzen2015terminator}, which involves the transport of two coupled tracers $X$ and $X_2$ on the surface of a sphere.
The chemicals interact such that the sum $m_T=m_X+2m_{X_2}$ should be preserved, where $m_X$ and $m_{X_2}$ are the mixing ratios of the individual chemicals.
When positivity or monotonicity limiters are applied to the numerical transport scheme, this may break the correlation between the two species which will be highlighted by this test.
This test is also a demonstration of simple ``physics-dynamics coupling'', as the conversion term between the species can be handled separately from the transport terms. \\
\\
For this test, we use the DG upwind scheme described in Section \ref{sec:transport} to transport the species on an icosahedral sphere mesh, and apply the vertex-based limiter of \cite{kuzmin2010vertex}.
The term describing the chemical reaction is separated from the transport term using a simple sequential splitting, \textcolor{red}{as described in Section \?}. \\
\\
Figure \ref{fig:terminator} shows the mixing ratios of the two individual species at the initial time and following a single rotation around the sphere by a deformational flow.
The change in the sum of the species is also displayed, with differences from zero remaining close to machine precision, which demonstrates that this transport scheme does preserve the correlation between the tracers.

\begin{figure}[htp!]
\centering
\includegraphics[width=0.9\textwidth]{terminator_toy.png}
\caption{The evolution of the chemical species from the ``Terminator Toy'' test of \cite{lauritzen2015terminator}.
The fields at the initial time are shown on the left, and at the final time ($t=12$ days) which follows a single revolution around the sphere by a deformational flow.
The bottom row shows the change in the total $m_T=m_X+2m_{X_2}$ from the initial time, which remains very close to zero.}
\label{fig:terminator}
\end{figure}

\subsection{Shallow water}
The standard suite of shallow water test cases originally proposed in
\citet{williamson1992standard} provides a common starting point for
testing new discretisations. This test suite is often augmented by
the unstable barotropic jet described in
\citet{galewsky2004initial}. \\
\\
Here we present two sets of convergence results for a variety of configurations measured using the steady state geostrophic balanced flow (test case 2 from
\citet{williamson1992standard}) with the linear shallow water equations. \JScomment{I think Hilary should also be cited here as she uses the linear version}
Figure \ref{fig:sw_convergence} shows the spatial convergence for a number of choices of finite element for the prognostic variables, using a Trapezium Rule discretisation (also known as Crank-Nicolson).
It also shows the temporal convergence for a range of timestepping schemes using the 1st order Raviart-Thomas elements on quadrilateral cells. 
\textcolor{red}{Once the test has been completed, come back to edit what we say here about convergence!} \\
\\
We show results from the unstable jet of \citet{galewsky2004initial} in Figure \ref{fig:galewsky}, using the non-linear shallow water equations.
These results use the Semi-Implicit Quasi-Newton timestepper described in Section \ref{???}, and compare the RT elements on a cubed-sphere mesh with the BDM elements on an icosahedral-sphere. \JScomment{both first order?}

\begin{figure}[htp!]
\centering
\includegraphics[width=0.9\textwidth]{}
\caption{Convergence results using the linear shallow water equations with the steady geostrophic flow of \cite{williamson1992standard}.
(Left) the spatial convergence of a number of orders and families of finite element, all using the same timestepper. (Right) the temporal convergence of different timesteppers, all using the same spatial discretisation.}
\label{fig:sw_convergence}
\end{figure}

\begin{figure}[htp!]
\centering
\includegraphics[width=0.9\textwidth]{galewsky_jet.png}
\caption{A view of the relative vorticity around the North pole at day 6 of a simulation of the Galewsky jet test case. (Left) the RT1 elements on a cubed-sphere mesh. (Right) the BDM1 elements on an icosahedral-sphere mesh. Both cases use the Semi-Implicit Quasi-Newton timestepper. The edges of the panels of the cube and icosahedron are highlighted.}
\label{fig:galewsky}
\end{figure}

\subsection{Compressible Euler equations}
We have also implemented many of the standard test cases for the non-hydrostatic compressible Euler equations that are used in modern NWP \JScomment{I think I've been saying `weather and climate' instead of NWP - happy to change but this is a note to remind us to be consistent!'} models.
In this section we show results from a small selection of these, which demonstrate different domains in which these methods work.
Throughout this section we use the Semi-Implicit Quasi-Newton timestepper.
 \\
\\
The first test is the falling density current of \citet{straka1993numerical} in a vertical slice. This test involves the addition of diffusion to both the potential temperature and velocity equations.
Figure \ref{fig:straka} shows the final potential temperature equation for simulations using the lowest-order and next-to-lowest order finite elements on the vertical slice.
The lowest-order methods require the transport and diffusion terms to use the ``recovered'' scheme of \citet{bendall2019recovered} to attain second-order accuracy. 
\\
\begin{figure}[htp!]
\centering
\includegraphics[width=0.8\textwidth]{straka.png}
\caption{The potential temperature field after 900 s from the falling density current test of \cite{straka1993numerical}. (Top) with the lowest-order finite elements, and (bottom) with the next-to-lowest-order finite elements.
}
\label{fig:straka}
\end{figure}
\\
To demonstrate the use of moist physics parametrisations, we use a moist three-dimensional rising thermal simulation in an extruded doubly-periodic plane.
The initial conditions and moisture parametrisation are those of \citet{bendall2020compatible}, which is an extension of the dry case of \citet{kelly2012continuous} and \citet{melvin2019mixed}.
A cross-sections of the final equivalent potential temperature field is shown in Figure \ref{fig:3d_bubble}.
As in the two-dimensional test of \citet{bryan2002benchmark}, the atmosphere is saturated and cloudy everywhere, so that the moist processes are active in the whole domain.
\\
\begin{figure}[htp!]
\centering
\includegraphics[width=0.48\textwidth]{moist_3d_bubble.png}
\caption{Vertical cross-section of the final equivalent potential temperature field from the three-dimensional moist rising thermal test.
}
\label{fig:3d_bubble}
\end{figure}
\\
Our final test case is the deep baroclinic wave of \citet{ullrich2014proposed} on the sphere.
An initial state in thermal wind balance is perturbed to produce a mid-latitude baroclinic instability, providing a flow that is typical of those found in Numerical Weather Prediction \JScomment{not sure I approve of these capitals...}.
The surface pressure and temperature fields after 10 days are shown in Figure \ref{fig:baroclinic}, which compare well with those in \citet{melvin2024mixed} and demonstrate that our methods can replicate results from the literature.

\begin{figure}[htp!]
\centering
\includegraphics[width=0.9\textwidth]{galewsky_jet.png}
\caption{The surface pressure and temperature after 10 days of a deep baroclinic wave simulation on the cubed-sphere.}
\label{fig:baroclinic}
\end{figure}
